---
title: "MovieLens Capstone Project"
author: "Akram Zalloum"
date: "Dec 2021"
output: pdf_document
---

## Introduction:
In this project we will explore the data set of MovieLens that has 10M movie ratings for 10K movies rated by 72K users. The dataset also include other columns like time, genere in addition to user information.

We need to predict ratings for movies based on the data we have training 90% and validation data 10%.

```{r Creat test and validation sets, echo=FALSE, message=FALSE, warning=FALSE}
##########################################################
# Create edx set, validation set (final hold-out test set)
##########################################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 3.6 or earlier:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                            title = as.character(title),
                                            genres = as.character(genres))
# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                            title = as.character(title),
                                            genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
      semi_join(edx, by = "movieId") %>%
      semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

```{r Loading packages, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(ggplot2)
```

## Understand data 
Based on the data structure we have 6 columns (userId,movieID,rating,timestamp,title,genres)
```{r}
str(edx)
summary(edx)
```
## Regarding the ratings statistics we have Min = 1, Max = 5, Mean = 3.51 and Mode = 4.0.
```{r echo=FALSE}
edx %>% group_by(rating) %>% summarize(count = n()) %>% top_n(5) %>%
	arrange(desc(count))  
```
## To get number of unique movies and users, we use the code:
```{r echo= FALSE}
edx %>% 
  summarize(n_users = n_distinct(userId),
            n_movies = n_distinct(movieId))
```

## The histogram below to get number of ratings by movie 

```{r  echo=FALSE}
edx %>% count(movieId) %>% ggplot(aes(n))+
  geom_histogram(color = "black" , fill= "light blue",bins = 30 , binwidth = 0.2)+
  scale_x_log10()+
  ggtitle(" Number of Rating per Movie")+
  theme_gray()
```

## To get number of ratings by user 

```{r echo= FALSE }
edx %>% count(userId) %>% ggplot(aes(n))+
  geom_histogram(color = "black" , fill= "light blue" , bins = 30, binwidth = 0.2)+
  ggtitle(" Number of Rating Per User")+
  scale_x_log10()+
  theme_gray()
```
 
## Visulaization below to show the rating by movie genre 

```{r echo=FALSE }
edx %>% separate_rows(genres, sep = "\\|") %>%
  group_by(genres) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>% ggplot(aes(genres,count)) + 
  geom_bar(aes(fill =genres),stat = "identity")+ 
  labs(title = " Number of Rating for Each Genre")+
  theme(axis.text.x  = element_text(angle= 90, vjust = 50 ))+
  theme_light()
 
```

## Top 10 genre 

```{r echo= FALSE }
edx %>% separate_rows(genres, sep = "\\|") %>%
	group_by(genres) %>%
	summarize(count = n()) %>%
	arrange(desc(count))
```
  
  
## Prepare training and testing data sets (80% to 20%)
```{r}
set.seed(1)
test_index <- createDataPartition(y = edx$rating, times = 1, p = 0.2, list = FALSE)
train_set <- edx[-test_index,]
test_set <- edx[test_index,]
```


## Calculate RMSE
```{r}
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2, na.rm = TRUE))
}
```

### Model 1
Predict the same rating for all movies regardless of the user.

```{r}
Mu_1 <- mean(train_set$rating)
Mu_1
```

```{r}
naive_rmse <- RMSE(test_set$rating,Mu_1)
naive_rmse
```

Get the RMSE and compare results.
```{r}
rmse_results <- data_frame(method = "Just the average", RMSE = naive_rmse)
rmse_results%>% knitr::kable()
```

### Model 2
 Where some movies rated more than others and based on the previous model will add $b~i$ to represent the average ranking for movie $i$ to show the biase, so we compute the average.
```{r}
Mu_2 <- mean(train_set$rating) 
movie_avgs <- train_set %>% 
  group_by(movieId) %>% 
  summarize(b_i = mean(rating - Mu_2))
```

Visulize results
```{r echo=FALSE}
movie_avgs %>% qplot(b_i, geom ="histogram", bins = 10, data = ., color = I("black"))
```

Prediction enhancement:
```{r}
predicted_ratings <- Mu_2 + test_set %>% 
  left_join(movie_avgs, by='movieId') %>%
  pull(b_i)
model_2_rmse <- RMSE(predicted_ratings, test_set$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie Effect Model",  
                                     RMSE = model_2_rmse))
rmse_results %>% knitr::kable()
                
```

### Model 3
Users who rate too many movies (>100)

```{r echo= FALSE}
train_set %>% 
  group_by(userId) %>% 
  summarize(b_u = mean(rating)) %>% 
  filter(n()>=100) %>%
  ggplot(aes(b_u)) + 
  geom_histogram(bins = 30, color = "black")
```

Based on the clear gaps between ratings will get the mean
```{r}
user_avgs <- train_set %>% 
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - Mu_2 - b_i))
  
```
Calculate RMSE again to see the enhancement 
```{r}
predicted_ratings <- test_set %>% 
  left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>%
  mutate(pred = Mu_2 + b_i + b_u) %>%
  pull(pred)
model_3_rmse <- RMSE(predicted_ratings, test_set$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie + User Effects Model",  
                                     RMSE = model_3_rmse))
rmse_results%>% knitr::kable()
```

## Apply RMSE on testing set

```{r}
valid_pred_rating <- validation %>%
  left_join(movie_avgs , by = "movieId" ) %>% 
  left_join(user_avgs , by = "userId") %>%
  mutate(pred = Mu_2 + b_i + b_u ) %>%
  pull(pred)
model_3_valid <- RMSE(validation$rating, valid_pred_rating)
rmse_results <-  bind_rows( rmse_results, data_frame(Method = "Validation Results" , RMSE = model_3_valid))
rmse_results%>% knitr::kable()
```

## Finding
Based on the methodology followed which is naive the 3rd model has best RMSE, after using the linear regression the findings that old movies has better prediction
